Effects of epidemiological structure on the transient evolution of HIV virulence

### 2016-10-25 ms/plosHIVcover_resubm.tex
cover letter; ms tweaks; fig tweaks

\revcomment{The abstract really needs to be improved. It’s missing the main points
of the paper (that different contact patterns change the predicted
trajectories and optimums of virulence), and has unnecessary minor
details (like the fact that parameters were sampled using a Latin
hypercube).}

\response{We've followed the reviewer's suggestions in adding
important points and removing unimportant details from the abstract}

\revcomment{The introduction is really lacking a cogent description of the study of
the evolution of virulence, and the motivation of the paper would be
totally unclear to someone not already very familiar with this field.
The main points that are needed to spell out this issue are:

[... detailed suggestions omitted]
}

\response{We thank the reviewer for pointing this out --- we've 
been over this ground so often that we've come to take this basic
framework for granted!  We've 
}
The paper claims to add considerations of population structure to the
study of virulence but only focuses on the rules of partnership
formation, while assuming all individuals have identical behavior
patterns, and not on another major aspect of population structure, which
is the heterogeneity in the number of partners (or rates of partnership
change) among individuals. In the case of a static network, this would
be considered the degree heterogeneity. Many studies have shown that
sexual contact networks tend to me more on the scale-free spectrum, and
so the assumption of uniformity of behavior seems weak. This study would
be most impactful if on top of examining these switch patterns it
examined the influence of contact heterogeneity.

In Figure 3 it would be helpful to also include time to progression to
AIDS as a figure panel

Reviewer #2: Modeling studies have shown that HIV could evolve towards
immediate virulence where the trade-off between host survival and
transmission is optimized. The aim of this study is to investigate how
different assumptions about sexual partnership dynamics affect the
transient evolution of HIV virulence in a population. While this is a
clear objective, the study is rather technical and, in my opinion, does
not provide major insights or conclusions. My major concerns regarding
the approach of the study is given below. Furthermore, there are several
issues in terms of structure and the model description that should be
considered by the authors.

Approach:
The authors argue that they compare more “realistic” models (based on
ref. 14) to the relatively simple model by Shireff et al. (ref. 10). In
my view, all these models are quite simplistic in that they do not take
into account age-structure and heterogeneity in sexual behavior.
Essentially, all these models are random mixing models where every
individual can make contact with every other individual. The only
difference to the random mixing model by Shireff et al. is that
transmission can occur in ongoing partnerships and that some models
allow for concurrency. Arguably much more important aspects that affect
HIV epidemics and virulence evolution are the above-mentioned age- and
risk-structure. Furthermore, the authors use a single-stage disease
model and do not distinguish between differences in transmissibility
between the acute (primary) and chronic (asymptomatic) stage of
infections, which might influence HIV virulence evolution more than
differences in the sexual partnership dynamics. Hence, I am not
convinced that the authors use a “more realistic” model than the one by
vShireff et al.

Structure:
- The references to the equations are not always correct.
- Figure 1 is described in Materials and Methods before the model is
fully explained. Consequently, the parameter r cannot be understood.
- Tables and Figures often lack dimensions, i.e., is is unclear what the
axes represent. Table 1 is more or less useless without providing
dimensions for the various parameters.
- The author summary seems to provide more information about the results
than the abstract. However, the authors use SPVL and progression to AIDS
as two different proxies for virulence which can be confusing.

Model:
The authors make use of a previously published model from ref. 14. This
model is based on the pair-formation formalism initially developed by
Dietz & Hadeler (1988, J Math Biol, PMID: 3351391) and later used by
Kretzschmar et al. (1994 & 1998, Math Biosci, PMID: 7833594 & 9597826).
I feel that the authors should refer to some of these original
publications as well. Unfortunately, the description of the models in
appendix S1 seems to be partially erroneous or incorrect. For example,
the primes for the derivatives are often missing (e.g., equation 22, 24
and 25) or it is unclear whether equations represent state variables or
derivates thereof. Also, the reasoning behind the “instswitch” models is
unclear to me. The authors say that “Once individuals leave a
partnership, they enter temporary compartments…”. The variables (or
derivatives?) X and Y_i in equation 19 consist of positive terms only,
so how do individuals leave these compartments? In my understanding the
standard pair model (equation 9) can easily account for instantaneous
partnership formation by simply setting the pair formation rate rho to a
high value.
 ------------------------------------------------------------------------

*Reviewers who wish to reveal their identities to the authors and other
reviewers should include their name here (optional). These names will
not be published with the manuscript, should it be accepted.*

Reviewer #1: (No Response)

Reviewer #2: (No Response)

>>>>

We appreciate the opportunity to revise our manuscript.
In addition to extensive clarification and correction following
the reviewers' comments, we implemented a more complex model
that includes variation in contact rates among individuals.

\hrule

\textbf{Reviewer \#1}
\revcomment{The abstract really needs to be improved. It’s missing the main points
of the paper (that different contact patterns change the predicted
trajectories and optimums of virulence), and has unnecessary minor
details (like the fact that parameters were sampled using a Latin
hypercube).}

\response{We've followed the reviewer's suggestions in adding
important points and removing unimportant details from the abstract.}

\revcomment{The introduction is really lacking a cogent description of the study of
the evolution of virulence, and the motivation of the paper would be
totally unclear to someone not already very familiar with this field.
The main points that are needed to spell out this issue are:
[... detailed suggestions omitted ...]
}

\response{We thank the reviewer for pointing this out --- we've 
been over this ground so often that we've come to take this basic
framework for granted!  We've  extensively revised the introduction
along the lines suggested by the reviewer.}

\revcomment{
The paper claims to add considerations of population structure to the
study of virulence but only focuses on the rules of partnership
formation, while assuming all individuals have identical behavior
patterns, and not on another major aspect of population structure, which
is the heterogeneity in the number of partners (or rates of partnership
change) among individuals. In the case of a static network, this would
be considered the degree heterogeneity. Many studies have shown that
sexual contact networks tend to me more on the scale-free spectrum, and
so the assumption of uniformity of behavior seems weak. This study would
be most impactful if on top of examining these switch patterns it
examined the influence of contact heterogeneity.}

\response{We added a new model to our study that allows for 
discrete classes of heterogeneity 
in contact rate, with activity-weighted random mixing; this
is a standard approach for implementing contact heterogeneity
in a compartmental model. We discuss other avenues for
increasing epidemiological realism (e.g., an explicitly
heterosexual model with sex-specific transmission and
progression characteristics, as well as age-structured
mixing patterns) and comment that they will require a
switch from our current compartmental approach to an agent-based
framework.}

\revcomment{In Figure 3 it would be helpful to also include time to progression to
AIDS as a figure panel}

\response{Our figures 2-5 are now framed in terms of expected
progression time rather than SPVL, as we thought this scale
would generally be more interpretable (as before, the supplementary
material includes summary figures based on the other metrics, in this
case SPVL and transmission probability).}

\textbf{Reviewer \#2}
\revcomment{Modeling studies have shown that HIV could evolve towards
immediate virulence where the trade-off between host survival and
transmission is optimized. The aim of this study is to investigate how
different assumptions about sexual partnership dynamics affect the
transient evolution of HIV virulence in a population. While this is a
clear objective, the study is rather technical and, in my opinion, does
not provide major insights or conclusions.
}

\response{Importance is in the eye of the beholder, of course; we
have tried to clarify the take-home points of our study
(epidemiological structures modify quantitative conclusions about
virulence evolution; realistic models that account for both
pair formation and extra-pair contact are best approximated by
random-mixing models and worst approximated by models with
intermediate complexity; extra-pair and other forms of unstructured
contact generally act to increase the rate of virulence evolution).
}

\revcomment{
The authors argue that they compare more “realistic” models (based on
ref. 14) to the relatively simple model by Shireff et al. (ref. 10). In
my view, all these models are quite simplistic in that they do not take
into account age-structure and heterogeneity in sexual behavior.
Essentially, all these models are random mixing models where every
individual can make contact with every other individual. The only
difference to the random mixing model by Shireff et al. is that
transmission can occur in ongoing partnerships and that some models
allow for concurrency. 
Arguably much more important aspects that affect
HIV epidemics and virulence evolution are the above-mentioned age- and
risk-structure.}

\response{As stated above in the response to reviewer \#1, we have
added a model with contact heterogeneity to our study; while
important, age-structured mixing is too complex for our current
modeling framework and will have to be left for future studies.}

\revcomment{
 Furthermore, the authors use a single-stage disease
model and do not distinguish between differences in transmissibility
between the acute (primary) and chronic (asymptomatic) stage of
infections, which might influence HIV virulence evolution more than
differences in the sexual partnership dynamics. Hence, I am not
convinced that the authors use a “more realistic” model than the one by
Shireff et al.}

\response{We agree that we have compromised the complexity of
the disease life-history model somewhat in order to explore 
the effects of epidemiological complexity.
Figure 1 represents our effort to show that
this simplification does not have a large impact on the results
(although it's always possible that there could be an interaction
between disease life-history complexity and epidemiological complexity
\ldots); we have tried to clarify this point (see ll. 84ff.).

\textbf{Structure}

\begin{itemize}
\item \revcomment{The references to the equations are not always correct.}
\response{We have corrected all the errors we could find.}
\item \revcomment{Figure 1 is described in Materials and Methods before the model is fully explained. Consequently, the parameter r cannot be understood.}
\response{We have added some detail to the caption.}
\item \revcomment{Tables and Figures often lack dimensions, i.e., is is unclear what the axes represent. Table 1 is more or less useless without providing
dimensions for the various parameters.}
\response{We have added dimensions throughout.}
\item \revcomment{The author summary seems to provide more information about the results than the abstract. However, the authors use SPVL and progression to AIDS as two different proxies for virulence which can be confusing.}
\response{We have thoroughly revised the abstract in response to reviewer \#1's comments. Commenting on both SPVL and progression time is admittedly tricky; in our revised version we have focused on SPVL only where making explicit comparisons with Shirreff \emph{et al.}'s models, and have otherwise stated results in terms of progression time.}
\end{itemize}

\textbf{Model}

\begin{itemize}
\item \revcomment{The authors make use of a previously published model from ref. 14. This
model is based on the pair-formation formalism initially developed by
Dietz \& Hadeler (1988, J Math Biol, PMID: 3351391) and later used by
Kretzschmar et al. (1994 \& 1998, Math Biosci, PMID: 7833594 \& 9597826).
I feel that the authors should refer to some of these original
publications as well.}
\response{We have added citations to these references.}

\item \revcomment{Unfortunately, the description of the models in
appendix S1 seems to be partially erroneous or incorrect. For example,
the primes for the derivatives are often missing (e.g., equation 22, 24
and 25) or it is unclear whether equations represent state variables or
derivates thereof.}
\response{We have fixed these errors.}
}
\item \revcomment{ Also, the reasoning behind the “instswitch” models is
unclear to me. The authors say that “Once individuals leave a
partnership, they enter temporary compartments…”. The variables (or
derivatives?) $X$ and $Y_i$ in equation 19 consist of positive terms only,
so how do individuals leave these compartments?}
\response{We have clarified that these terms are partnership
leaving rates, and have added several lines of text clarifying
the meaning of the term.}
\item \revcomment{
 In my understanding the
standard pair model (equation 9) can easily account for instantaneous
partnership formation by simply setting the pair formation rate rho to a
high value}
\response{We have added the following sentence (ll. 141-145):
``Although these models can also be implemented
by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.''}

\end{itemize}

### 2016-10-25 ms/plos_latex.tex
cover letter; ms tweaks; fig tweaks

\end{flushleft}
% Please keep the abstract below 300 words
% current is 282
\section*{Abstract}
The evolutionary dynamics of parasite virulence change
in important ways over the
course of an emerging epidemic.
Changes in the fitness landscape
generally select for higher virulence early in an
epidemic; however, quantitative outcomes
may depend sensitively on epidemiological details and the structure of
mathematical models used to portray them.  Fraser \emph{et al.} 
have proposed a model for the eco-evolutionary
dynamics of HIV that captures
the tradeoffs between transmission and virulence (mediated by
set-point viral load, SPVL) and their heritability between
hosts. However, these models use implicit
representations of the transmission process that drastically simplify the
partnership dynamics that previous research has found to be critical
in driving epidemics of sexually transmitted diseases.  
Our models combine HIV virulence tradeoffs with a range of
epidemiological structures, modeling partnership formation and
dissolution and allowing for individuals to transmit disease outside
of partnerships. We assess summary statistics such as the peak virulence
(corresponding to the minimum expected time of progression to AIDS) across all
models for a range of 
partnership dynamic parameters 
applicable to the HIV epidemic in sub-Saharan Africa.
Although virulence trajectories are broadly similar
across model structures, the timing and magnitude of the 
minimum expected time to progression vary
considerably.
Models of intermediate
complexity as used by Fraser \emph{et al.} \todo{why not shirreff et al?} 
predicted lower slower progression/lower virulence (a minimum of 15 years to
progress to AIDS) compared to both more realistic models
and simple random-mixing models with no partnership structure
at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
In this range of models, the simplest random-mixing structure best
approximates the most realistic model; this
surprising outcome occurs because the dominance of extra-pair
contact in the realistic model tends to swamp the effects of
partnership structure.

@@

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
disease strains that can reproduce more --- where reproduction
is defined here as the amount of \emph{between-host} transmission, or
the number of new hosts infected --- will increase in prevalence.
Pathogens can increase their net reproduction rate either
by increasing their transmission rate, 
the rate (per infected host) at which they
infect new hosts, or by decreasing their clearance or disease-induced
mortality rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
the transmission and disease-induced mortality rate are both linked to the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
Fraser \etal\ have applied these ideas in a particularly
interesting way 
by showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was both
heritable and covaried with the set-point viral load (i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection), which in turn predicted the probability of
transmission
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

Like Shirreff \etal\ \cite{shirreff_transmission_2011}, we focus on the evolution of mean $\log_{10}$ set-point viral load, SPVL (which we denote as $\alpha$), rather than the rate of progression to AIDS itself
(hereafter ``virulence'' will refer either to the SPVL, or to 
the rate of progression to AIDS; these two quantities are
deterministically linked in the model) \todo{Doesn't this somewhat contradict with what we say later on? Here, we say we focus on SPVL but later we say we summarize the models using progression time rather than SPVL...}.
In contrast to Shirreff \etal, we use a single-stage disease model instead of accounting explicitly for progression through the three main stages of HIV infection (primary, asymptomatic, and disease), and we use a simple exponentially distributed infectious period instead of a more realistic Weibull-distributed infectious period; we show below that our results are not overly
sensitive to this simplification. We account for varying transmission rates and durations of each disease stage by summing the durations of three stages (again based on Shirreff \etal's model) and taking the duration-weighted average of transmission rates of three stages. Thus the within-couple transmission rate, $\beta$, for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

@@

The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch''). The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$. Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission. \todo{Although these models can also be achieved by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.}

@@

Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). \todo{does this sentence regarding heterogeneity feel redundant?} For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well. We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

@@

Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity.
%%  BMB: I don't think we need to go into detail about variance vs CV
To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men, as that was the group with the largest variability \cite{omori2015dynamics}. 
Omori \etal \cite{omori2015dynamics} give a very wide
range for the coefficient of variation ($\approx$ 0 - 20, corresponding
to squared CV range of 0-400):
we narrowed this range for $\textrm{CV}^2$ to 0.01-100.
At the bottom end of the range, estimating that a group behaves
perfectly homogeneously ($CV=0$) is likely to be a sampling artifact;
at the upper end, the estimate is also likely to be noisy because
of the low mean value among married females (who have the largest
range of CV). 
We assume that the number of non-cohabiting partners follows a Gamma distribution.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$. After each run, initial exponential growth rate is calculated. Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
For calibration purposes, we run each model for only 500 years
(full simulations are run for 4000 years), which is always long
enough to capture the exponential growth phase of the model. 
We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations. \todo{mention that we got rid of error runs for hetero model?}

@@

Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$. Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - c II_{ij}$. Once individuals leave a partnership, they are instantaneously distributed into coupled states. In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the number of susceptible individuals that leave partnership at a given time, and $Y_i$ the number of infected individuals with \Lspvl\ of $\alpha_i$ who leave partnership at a given time. These temporarily single individuals then form couples through the same partnership formation rule described in the previous section:

@@

We extend the ``pairform+epc'' model by allowing for heterogeneity in sexual behaviour. Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group. In other words, an individual in a higher risk is also more likely to form a \todo{long-term sexual partnership instead of stable?} stable partnership, leave a stable partnership, and engage in a extra-couple/uncoupled mixing. We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group. For simplicity, we assume that the transmission rate per partnership is unaffected by sexual behaviour.

@@

For dissolution process, an individual in risk group $i$ leaves its partnership at a rate $\varphi_i c$. If a partnership is formed between two individuals from a different risk group, the rates at which they leave the partnership differ. We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners. Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate. \todo{Herbeck 2016 does this. We can cite them?}

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
% current is 282
\section*{Abstract}
The evolutionary dynamics of parasite virulence change
in important ways over the
course of an emerging epidemic.
Changes in the fitness landscape
generally select for higher virulence early in an
epidemic; however, quantitative outcomes
may depend sensitively on epidemiological details and the structure of
mathematical models used to portray them.  Fraser \emph{et al.} 
have proposed a model for the eco-evolutionary
dynamics of HIV that captures
the tradeoffs between transmission and virulence (mediated by
set-point viral load, SPVL) and their heritability between
hosts. However, these models use implicit
representations of the transmission process that drastically simplify the
partnership dynamics that previous research has found to be critical
in driving epidemics of sexually transmitted diseases.  
Our models combine HIV virulence tradeoffs with a range of
epidemiological structures, modeling partnership formation and
dissolution and allowing for individuals to transmit disease outside
of partnerships. We assess summary statistics such as the peak virulence
(corresponding to the minimum expected time of progression to AIDS) across all
models for a range of 
partnership dynamic parameters 
applicable to the HIV epidemic in sub-Saharan Africa.
Although virulence trajectories are broadly similar
across model structures, the timing and magnitude of the 
minimum expected time to progression vary
considerably.
Models of intermediate
complexity as used by Shirreff \emph{et al.}
predicted lower slower progression/lower virulence (a minimum of 15 years to
progress to AIDS) compared to both more realistic models
and simple random-mixing models with no partnership structure
at all (both with a minimum of $\approx$ 7.25 years to progress to AIDS).
In this range of models, the simplest random-mixing structure best
approximates the most realistic model; this
surprising outcome occurs because the dominance of extra-pair
contact in the realistic model tends to swamp the effects of
partnership structure.

@@

The evolution of pathogen virulence has both 
theoretical and, potentially, practical
importance. In general, evolutionary theory suggests that
disease strains that can reproduce more --- where reproduction
is defined here as the amount of \emph{between-host} transmission, or
the number of new hosts infected --- will increase in prevalence.
Pathogens can increase their net reproduction rate either
by increasing their transmission rate, 
the rate (per infected host) at which they
infect new hosts, or by decreasing their clearance or disease-induced
mortality rate, the rate
at which hosts recover or die from disease.
The \emph{trade-off theory} \cite{alizon_virulence_2009} postulates that
the transmission and disease-induced mortality rate are both linked to the rate at which
the pathogen exploits host resources for within-host reproduction, 
and that pathogen
evolution will thus strike a balance between the
pathogen's rate of transmission
to new hosts and its rate of killing its host (or of provoking
the host's immune system to eliminate it).
Some biologists have criticized the tradeoff theory
\cite{EbertBull2003,alizon_adaptive_2015}, but others have
successfully applied it to a variety of host-pathogen systems \cite{Dwyer+1990,mackinnon1999genetic,jensen2006empirical,deroode2008virulence}.
Fraser \etal\ have applied these ideas in a particularly
interesting way 
by showing that HIV appears to satisfy the prerequisites of
the tradeoff theory: in studies of discordant couples (i.e. long-term
sexual partnerships with one infected and one uninfected partner), HIV
virulence as measured by the rate of progression to AIDS was 
both heritable and covaried with the set-point viral load (SPVL: i.e., the
characteristic virus load measured in blood during the intermediate
stage of infection) and the probability of transmission.
Higher viral loads led to shorter progression times (higher virulence)
and faster transmission, with a decelerating relationship
between virulence and transmission as required by the tradeoff theory
\cite{Fraser+2007,fraser_virulence_2014}. Subsequent studies
\cite{shirreff_transmission_2011,herbeck_hiv_2014,herbeck_evolution_2016} used these data to
parameterize mechanistic models of HIV virulence evolution, suggesting
that HIV invading a novel population would initially evolve increased
virulence, peaking after approximately 100-200 years and then declining
slightly to a long-stable virulence level.

@@

As in Shirreff \etal\ \cite{shirreff_transmission_2011}, our models
explicitly track the evolution of mean $\log_{10}$ SPVL
(which we denote as $\alpha$), rather than the rate of progression to
AIDS itself (hereafter ``virulence'' will refer either to the SPVL, or
to the rate of progression to AIDS; these two quantities are
deterministically linked in the model).  
In contrast to Shirreff \etal, we use a
single-stage disease model instead of accounting explicitly for
progression through the three main stages of HIV infection (primary,
asymptomatic, and disease), and we use a simple exponentially
distributed infectious period instead of a more realistic
Weibull-distributed infectious period; we show below that our results
are not overly sensitive to this simplification. We account for
varying transmission rates and durations of each disease stage by
summing the durations of three stages (again based on Shirreff \etal's
model) and taking the duration-weighted average of transmission rates
of three stages. Thus the within-couple transmission rate, $\beta$,
for our models is given by:
\begin{equation}
\beta (\alpha) = \frac{D_P \beta_P + D_A (\alpha) \beta_A (\alpha) + D_D \beta_D}{D_P + D_A (\alpha) + D_D},
\end{equation}
where the duration of infection ($D_P$ and $D_D$) and rate of transmission ($\beta_P$ and $\beta_D$) of the Primary and Disease stages
of infection are independent of the host's SPVL. Following Shirreff \etal, the duration of infection ($D_A$) and rate of transmission ($\beta_A$) for the Asymptomatic stage are Hill functions of the SPVL:

@@

The next two models, which are intended to bridge the gap between models with fully explicit pair-formation dynamics and the simpler, implicit models used by Shirreff \emph{et al.} \cite{shirreff_transmission_2011}, assume instantaneous partnership formation (``instswitch''). The compartmental structure thus omits the single states $S$ and $I$, comprising only the three partnered states: $SS$, $SI$, and $II$. Like the first two models, this pair of models differs in their inclusion of extra-pair contact: the third model (``instswitch+epc'') includes extra-partnership contact (now only with individuals in other partnerships, since uncoupled individuals do not exist in this model) while the fourth (``instswitch'') only considers within-couple transmission. Although these models can also be implemented
by setting the partnership formation rate of the explicit partnership models to a high value (and we have tested that both methods in fact produce same results), we model instantaneous partnership formation models independently in order to avoid scaling of partnership formation rate during model calibration affecting the virulence trajectory.

@@

Despite considerable effort \cite{hollingsworth_hiv1_2008,champredon_hiv_2013}, the parameters determining the rate and structure of sexual partnership change and contact are still very uncertain; this led Champredon \etal\ \cite{champredon_hiv_2013} to adopt a Latin hypercube sampling (LHS) strategy \cite{blower_drugs_1991} that evaluates model outcomes over a range of parameter values. In order to make sure that our comparisons among models apply across the entire space of reasonable parameter values, and in order to evaluate the differential sensitivity of different model structures to parameter values, we follow a similar protocol and perform LHS over a parameter set including both the early- and late-stage transmission and duration parameters ($\beta_P$, $D_P$, $\beta_D$, $D_D$) and contact/partnership parameters ($\rho$, $c$, $c_u/c_w$, and $c_e/c_w$). For the heterogeneity model, the mean and squared coefficient of variation (CV) for the number of non-cohabiting partners are sampled as well. We do not allow for uncertainties in parameters that are directly related to the evolutionary process ($\tsub{\beta}{max}$, $\beta_{50}$, $\beta_k$, $\tsub{D}{max}$, $D_{50}$, $D_k$, $\sigma_M$), instead using Shirreff \etal's point estimates throughout \cite{shirreff_transmission_2011}.

@@

Two parameters, mean and the squared coefficient of variation (CV) of number of non-cohabiting partners, are sampled for heterogeneity in sexual activity.
To allow for a wide range of uncertainty, range for the mean number of non-cohabiting partners was taken from unmarried men, as that was the group with the largest variability \cite{omori2015dynamics}. 
Omori \etal \cite{omori2015dynamics} give a very wide
range for the coefficient of variation ($\approx$ 0 - 20, corresponding
to squared CV range of 0-400):
we narrowed this range for $\textrm{CV}^2$ to 0.01-100.
At the bottom end of the range, estimating that a group behaves
perfectly homogeneously ($CV=0$) is likely to be a sampling artifact;
at the upper end, the estimate is also likely to be noisy because
of the low mean value among married females (who have the largest
range of CV). 
We assume that the number of non-cohabiting partners follows a Gamma distribution.

@@

We run each model for each of 1000 parameter sets chosen by Latin hypercube sampling, with fixed starting conditions
of mean \Lspvl\ of 3 and epidemic size of $10^{-4}$. After each run, initial exponential growth rate is calculated. Then, parameters are scaled so that the initial exponential growth rate is scaled to 0.04, a value that approximates the growth rates of Shirreff \etal's original models.
For calibration purposes, we run each model for only 500 years
(full simulations are run for 4000 years), which is always long
enough to capture the exponential growth phase of the model. 
We use a 4/5 order 
Runge-Kutta method (\texttt{ode45} from the \texttt{deSolve} package
\cite{soetaert_solving_2010}) for all simulations. 
(For the heterogeneous model, approximately
10\% of the samples failed due to numerical instability.)

@@

Since model 3 and 4 assume instantaneous partnership formation, there are only three states: $SS$, $SI_i$, and $II_{ij}$. Partnership dissolution rates are equal to those of model 1 and 2: $\DS(SS) = -cSS$, $\DS(SI_i) = - cSI_i$, and $\DS(II_{ij}) = - c II_{ij}$. Once individuals leave a partnership, they are instantaneously distributed into coupled states. In order to make the equations simpler, we introduce the following two terms: $X$ and $Y_i$, where $X$ denotes the total rate at which susceptible individuals leave partnership at a given time, and $Y_i$ is the total partnership leaving rate of infected individuals with \Lspvl\ of $\alpha_i$ at a given time. These temporarily single individuals then form couples through the same partnership formation rules described in the previous section:

@@

We extend the ``pairform+epc'' model by allowing for heterogeneity in sexual behaviour. Since pairform+epc model captures four distinct sexual behaviours -- pair formation, pair dissolution, extra-couple mixing, and uncoupled mixing -- we assume that all four parameters that model these behaviours are scaled by the same factor based on the risk group. In other words, an individual in a higher risk is also more likely to form a long-term sexual partnership, leave a long-term partnership, and engage in a extra-couple/uncoupled mixing. We denote this scaling parameter as $\varphi_i$ where $i$ is the risk group. For simplicity, we assume that the transmission rate per partnership is unaffected by sexual behaviour.

@@

For dissolution process, an individual in risk group $i$ leaves its partnership at a rate $\varphi_i c$. If a partnership is formed between two individuals from a different risk group, the rates at which they leave the partnership differ. We resolve this conflict by assuming that a partnership dissolution rate of a couple is equal to the average of that of two partners. Therefore, $XY_{ij, kl}$ dissolve at per-partnership rate $(\varphi_k + \varphi_l)/2 c$, and both $X_{i,k}$ and $Y_{j,l}$ partners return to single state at the same rate \cite{herbeck_evolution_2016}.
